Solar wind and geomagnetic activity influence on the equatorial plasma mass density at geosynchronous orbit

### 2016-06-28 paper/paper.tex
Misc. edits

\twocolumn[
  \begin{@twocolumnfalse}
\maketitle
\hrule
\begin{abstract}
In this work we consider two types of events, determined by decreases in $D_{st}$ below a threshold value or increases in the $\rho_{eq}$ above a threshold value on 1-hour and 1-day time scales using the \cite{Denton2007} dataset.  From the $D_{st}$ events, we find that there is no statistically significant relationship between either solar wind or geomagnetic activity under conditions of strong geomagnetic activity when hourly or daily averages of $\rho_{eq}$ are considered.  Although very large geomagnetic storms have been observed to correspond to large increases in the plasma trough mass density, in general $\rho_{eq}$ following the onset of a geomagnetic storm does not depend on the storm intensity or the solar wind parameters $B_{z}$ and $P_{dyn}$. From the $\rho_{eq}$ events, we find a weak dependence on $B_{z}$ after the onset of an event, with higher average $B_{z}$ six hours after the event onset corresponding to larger $\rho_{eq}$ events.
\end{abstract}
\vspace{2em}
\hrule
\vspace{2em}
 \end{@twocolumnfalse}
  ]

@@

The dependence of the plasma trough density on geomagnetic and solar wind conditions has not been extensively studied until the past decade (\cite{Takahashi2006}; \cite{Takahashi2010}; \cite{Denton2016}). Earlier works that considered profiles of electron number density found a significant dependence on the plasmasphere boundary on geomagnetic activity (\cite{Chappell1970}; \cite{Anderson1993}), but there the relationship between geomagnetic activity and density in the plasma trough was not explicitly considered.

\cite{Takahashi2006} estimated magnetospheric mass density using measurements from the CRRES satellite during a 73-day period in 1991 (observations were made when CRRESS was in an elliptical, near equatorial orbit and primarily in the plasma trough region). They found that the average ion mass, $M=\rho/n_e$, had some correlation with geomagnetic activity, with more negative hourly $D_{st}$ values corresponding to higher $M$ and 1.5- and 3-day averages of $K_p$ corresponding to higher $M$ (3-hour averages of $K_p$ had little visual correlation with $M$). The mass density estimates were obtained from CRRES observations in the MLT range of 12:00 and 18:00 with most observations having L between 5 and 7 $R_E$.

\cite{Denton2006} found a weak relationship between $D_{st}$ and and $K_p$ and the field line distribution of $\rho$ using observations of the first three toroidal Alfv\'en harmonic frequencies for $L$ in the range of 6-8$R_E$.  A more pronounced peak in the distribution along a filed line near the equator was found for lower $D_{st}$ and larger $K_p$.  The equatorial value of $\rho$ appeared identical for the bins of $D_{st}$ (-142 to -31 nT and -31 to 37 nT) and $K_p$ (1.5 to 3.4 and 3.4 to 5.9) considered .

\cite{Takahashi2010} developed a mass density dataset using measurements from the Space Environment Monitor instruments on the Geostationary Operational Environmental Satellites (GOES) satellites from 1980 through 1992, with most measurements in the range of $L=6.8\pm0.2 R_E$. The mass density $\rho$ was estimated using the the Alfv\'en wave velocity relationship, $V_A=B/\sqrt{\mu_0\rho}$ a magnetic field model, and a numerical solution to a wave equation with an ionospheric boundary and the assumption of a zero resistance ionosphere.  The equatorial mass density, $\rho_{eq}$, was derived from the estimated mass density using a power law dependence on the geocentric distance to the field line of the observation, $R$, $\rho=\rho_{eq}L/(R/R_E)$.  They found a high correlation ($\sim 0.93$) between 27-day averages of F10.7 and 27-day medians of $\rho_{eq}$.

@@

The parameters $\rho_{eq}$ and $F_{10.7}$ used in this work are from the dataset of \cite{Denton}, which was used in \cite{Takahashi2010}; data are available from 1980 through 1991 from GOES 2,3,5, 6 and 7.  All other parameters used are from \cite{Kondrashov2014ReconstructionOfGaps} from 1972 through 2013, which are on a 1--hour time grid. $\rho_{eq}$ is the inferred equatorial mass density based on the 3rd harmonic torodial frequency of magnetic field measurements as described in \cite{Takahashi2010}.  The smallest cadence for $\rho_{eq}$ values is 10 minutes.  To compute an hourly median over the same interval for which the solar wind parameters were averaged, the median of all $\rho_{eq}$ values in a given hour window was used.  Fill (NaN) values were used for hours in which no measurements were available.  In cases where $\rho_{eq}$ was available from multiple GOES satellites at the same time, the value from GOES 6 was selected.  In the dataset, measurements from GOES 6 had the longest time span of coverage.  An overview of the data used is shown in Figure~\ref{AllData}.

In this work, events are defined to occur when $D_{st}$ or $\rho_{eq}$ crosses a threshold value, as indicated by horizontal dashed lines in their respective panels of Figure~\ref{AllData}.

Figure~\ref{ccplot} shows the long-term trends of $log(\rho_{eq})$ from GOES 6 and $F_{10.7}$ computed using the median values in 27-day non-overlapping windows.  A scatter plot of these two lines is shown in the lower panel of Figure~\ref{ccplot}.  The linear correlation for GOES 6 using all measurements is found to be $0.94$, which is the same value documented by \cite{Takahashi2010} who used measurements from all satellites in the time interval of 1980 through 1991 using the constraint $0600 \leq MLT \le 1200$, $K_{p3d}\ge1.0$ and $D_{st}\ge -50$~nT.  Our 27-day correlations for GOES 2, 5, and 7 are 0.81, 0.78, and 0.87, and the time range of coverage of measurements from these satellites are 1980-01-01/1983-05-16, 1983-01-01/1987-02-28, and 1983-05-27/1991-08-29 respectively.

@@

Two types of events are considered. The first is a drop in the $D_{st}$ index \vinote{Cite Kyoto} below the threshold of -50~nT. These types of events are considered in order to determine if there is a well-defined $\rho_{eq}$ dependence on geomagnetic activity/storms as indicated by $D_{st}$ index.  It is expected that the plasmasphere will exhibit two responses near these threshold crossings.  First is the movement of the plasmasphere earthward (\cite{LemaireEarthsPlasmasphere}) and the second is a possible change in density due to magnetospheric processes.

The second type of event is an increase in $\rho_{eq}$ above a threshold of 20~amu/cm$^3$.  It is expected that these increases will typically occur after period of quiet geomagnetic activity. In this case, the plasmapause boundary may move outward past geosynchronous orbit and the increase will be due to measurements being made inside the plasmapause.  A second possible reason for the increase could be due to magnetospheric processes.

@@

Our first analysis uses data only from the time interval 1989-1991, which corresponds with the interval used in the analysis in \cite{Takahashi2010}. In this interval, there were 329 $D_{st}$ events and 70 $\rho_{eq}$ events.  For both types of events, the hour of the threshold crossing defines the zero epoch time, and for each event, $4\cdot24$ hours were considered before each event and $8\cdot24$ hours after each event.  To compute the epoch averages on a daily time scale, the median value of all available measurements for all events centered on a window of $\pm 12$ hours of the epoch zero hour was computed, and these averaging windows were shifted in increments of $24$~hours. Similar results are obtained if we first reduce each event time series to have a 1-day cadence by computing medians in 1-day bins and then compute the medians across events.  Error bars for each parameter are the standard deviation of the values used in computing the median divided by the square root of the number of values used.

The stack plot in the upper panel of Figure \ref{DailyAveragedDstEvents} shows the epoch averages for these events.  The minimum $D_{st}$ median is $-48$~nT.  Consistent with \cite{Takahashi2010}, $D_{st}$ events correspond to elevated $\rho_{eq}$, although the magnitude of increase observed here is $1.5$ amu/cm$^3$ (from 18 to 19.5~amu/cm$^3$) instead of the increase of $\sim 10$ amu/cm$^3$ found in Figure 11 of \cite{Takahashi2010}.  The vertical green bars in the $\rho_{eq}$ plot show the number of $\rho_{eq}$ values that were used to compute the medians and error bars.  For the non-\req\ medians, the number of measurements used in computing the medians is typically equal to the number of events, which is much larger than the number of \req\ values as we used all available measurements instead of restricting to only values where a non--fill $\rho_{eq}$ value existed.

In the lower panel of Figure \ref{DailyAveragedDstEvents} it is shown that the trends seen in the analysis of 1989-1991 also hold for the entire span of measurements from GOES 6, which spans 1983-1991 - a small elevation on the day of the event relative to the day before and day after, with median averages after the day of the event being slightly lower than that before.  
To determine if the magnitude of the density medians in the top panel of \ref{DailyAveragedDstEvents} on days -1, 0, and 1 are statistically different from each other, we used a two-population bootstrap test.  A bootstrap median was created by sampling the values used to compute the median on each day with replacement.  The observed difference between the medians is shown in the table along with the fraction of observations in 1000 bootstrap differences with a different sign than that observed.  The results are shown in the fourth column of Table~\ref{BootstrapDifferenceTable}. The fourth column shows the results of a 2-population t-test and the fifth column indicates ???.

@@

Because the dynamics of the magnetopshere take place on a time scale of less than one day, we have also considered hourly averages as shown in the top panel of Fig.~\ref{HourlyAveragedDstEvents}.  A total of 668 events were found in the interval of May 1983 through August 1991 with an average duration of 9 hours and a median duration of 3 hours.  The error bars were computed in the same way as Fig.~\ref{DailyAveragedDstEvents}.  In this plot, the overall trend is similar to that found in the bottom panel of Fig.~\ref{HourlyAveragedDstEvents} - $\rho_{eq}$ is slightly elevated at the time of the threshold crossing of $D_{st}$ with respect to the values one day before and after, but the elevation is small and in general not statistically significant.  At this timescale no obvious change in mass density at hourly timescale is visible. This points to an issue with only looking at long-timescale trends between density and $D_{st}$, and allows for the possibility that other factors are influencing the long term correlation because there is no obvious connection on a short timescale.  One possibility is that, as suggested in \cite{Takahashi2010}, $F_{10.7}$ plays a significant role in driving long term density values which biases the long-term correlation of density and $D_{st}$.

The bottom panel of Fig.~\ref{HourlyAveragedDstEvents} contains the events shown in the top panel under the constraint that $D_{st}$ remained below -50~nT for at least twelve hours after the event onset time.  In the time period of elevated mass density within 12 hours of the start of the event, there were an average of \inote{X} samples per hourly interval.  This result is consistent with previous observations of large mass density spikes in the plasma trough region during large geomagnetic storms (\cite{Yao2008}; \cite{Takahashi2010} and the observation that solar wind parameters have very little correlation independent of $F_{10.7}$ (\cite{Denton2016}).  To determine of this relationship can be observed in the events of the top panel of Fig.~\ref{HourlyAveragedDstEvents}, we sorted the events by the average $D_{st}$ values in the time interval of 0-12 hours, split the events in two and recomputed the epoch averages of $\rho_{eq}$.  The result is that there is no statistically significant difference in the mass density epoch averages for large versus small storms.  This indicates that to confirm the $\rho_{eq}$ dependence on large geomagnetic storms would require a dataset with a much larger number of strong geomagnetic storms.

>>>>

\twocolumn[
  \begin{@twocolumnfalse}
\maketitle
\hrule
\begin{abstract}
In this work we consider two types of events, determined by decreases in $D_{st}$ below a threshold value or increases in the the equatorial mass density at geosynchronous altitudes, $\rho_{eq}$, above a threshold value on 1-hour and 1-day time scales using the \cite{Takahashi2010} dataset.  From the $D_{st}$ events, we find that there is no statistically significant.  Although very large geomagnetic storms have been observed to correspond to large increases in the $\rho_{eq}$, in general $\rho_{eq}$ following the onset of a geomagnetic storm does not depend on the storm intensity or the solar wind parameters $B_{z}$ and $P_{dyn}$. From the $\rho_{eq}$ events, we find a weak dependence on $B_{z}$ after the onset of an event, with higher average $B_{z}$ six hours after the event onset corresponding to larger $\rho_{eq}$ events.
\end{abstract}
\vspace{2em}
\hrule
\vspace{2em}
 \end{@twocolumnfalse}
  ]

@@

The dependence of the plasma trough density on geomagnetic and solar wind conditions has not been extensively studied until the past decade (\cite{Takahashi2006}; \cite{Takahashi2010}; \cite{Denton2016}). Earlier works that considered profiles of hydrogen ion density found a significant dependence on the plasmasphere boundary on geomagnetic activity (\cite{Chappell1970}; \cite{Maynard1977}; \cite{Carpenter1992}), but there the relationship between geomagnetic activity and density in the plasma trough was not explicitly considered.

\cite{Takahashi2006} estimated magnetospheric mass density using measurements from the CRRES satellite during a 73-day period in 1991 (observations were made when CRRESS was in an elliptical, near equatorial orbit and primarily in the plasma trough region). They found that the average ion mass, $M=\rho/n_e$, had some correlation with geomagnetic activity, with more negative hourly $D_{st}$ values corresponding to higher $M$ and 1.5- and 3-day averages of $K_p$ corresponding to higher $M$ (3-hour averages of $K_p$ had little visual correlation with $M$). The mass density estimates were obtained from CRRES observations in the MLT range of 12:00 and 18:00 with most observations having $L$ between 5 and 7 $R_E$.

\cite{Denton2006} found a weak relationship between $D_{st}$ and and $K_p$ and the field line distribution of $\rho$ using observations of the first three toroidal Alfv\'en harmonic frequencies for $L$ in the range of 6-8~$R_E$.  A more pronounced peak in the distribution along a filed line near the equator was found for lower $D_{st}$ and larger $K_p$.  The equatorial value of $\rho$ appeared identical for the bins of $D_{st}$ (-142 to -31 nT and -31 to 37 nT) and $K_p$ (1.5 to 3.4 and 3.4 to 5.9) considered .

\cite{Takahashi2010} developed a mass density dataset using measurements from the Space Environment Monitor instruments on the Geostationary Operational Environmental Satellites (GOES) satellites from 1980 through 1992, with most measurements in the range of $L=6.8\pm0.2 R_E$. The mass density $\rho$ was estimated using the the Alfv\'en wave velocity relationship, $V_A=B/\sqrt{\mu_0\rho}$, a magnetic field model, and a numerical solution to a wave equation with an ionospheric boundary and the assumption of a zero resistance ionosphere.  The equatorial mass density, $\rho_{eq}$, was derived from the estimated mass density using a power law dependence on the geocentric distance to the field line of the observation, $R$, $\rho=\rho_{eq}L/(R/R_E)$.  They found a high correlation ($\sim 0.93$) between 27-day averages of F10.7 and 27-day medians of $\rho_{eq}$.

@@

The parameters $\rho_{eq}$ and $F_{10.7}$ are from the dataset of \cite{Denton}, which was used in \cite{Takahashi2010}; data are available from 1980 through 1991 from GOES 2, 3, 5, 6 and 7.  All other parameters used are from \cite{Kondrashov2014ReconstructionOfGaps} from 1972 through 2013, which are on a 1--hour time grid. $\rho_{eq}$ is the inferred equatorial mass density based on the 3rd harmonic torodial frequency of magnetic field measurements as described in \cite{Takahashi2010}.  The smallest cadence for $\rho_{eq}$ values is 10 minutes.  To compute an hourly median over the same interval for which the solar wind parameters were averaged, the median of all $\rho_{eq}$ values in a given hour window was used.  Fill (NaN) values were used for hours in which no measurements were available.  In cases where $\rho_{eq}$ was available from multiple GOES satellites at the same time, the value from GOES 6 was selected.  In the dataset, measurements from GOES 6 had the longest time span of coverage.  An overview of the data used is shown in Figure~\ref{AllData}.

Figure~\ref{ccplot} shows the long-term trends of log$_{10}(\rho_{eq})$ from GOES 6 and $F_{10.7}$ computed using the median values in 27-day non-overlapping windows.  A scatter plot of these two lines is shown in the lower panel of Figure~\ref{ccplot}.  The linear correlation for GOES 6 using all measurements is found to be $0.94$, which is the same value documented by \cite{Takahashi2010} who used measurements from all satellites in the time interval of 1980 through 1991 using the constraint $0600 \leq MLT \le 1200$, $K_{p3d}\ge1.0$ and $D_{st}\ge -50$~nT.  Our 27-day correlations for GOES 2, 5, and 7 are 0.81, 0.78, and 0.87, and the time range of coverage of measurements from these satellites are 1980-01-01/1983-05-16, 1983-01-01/1987-02-28, and 1983-05-27/1991-08-29 respectively.

@@

Two types of events are considered. The first is a drop in the $D_{st}$ index below the threshold of -50~nT. These types of events are considered in order to determine if there is a well-defined $\rho_{eq}$ dependence on geomagnetic activity/storms as indicated by $D_{st}$ index.  It is expected that the plasmasphere will exhibit two responses near these threshold crossings.  First is the movement of the plasmasphere earthward (\cite{LemaireEarthsPlasmasphere}) and the second is a possible change in density due to magnetospheric processes.

The second type of event is an increase in $\rho_{eq}$ above a threshold of 20~amu/cm$^3$.  It is expected that these increases will typically occur after period of quiet geomagnetic activity but may also be the result of increases in geomagnetic activity. In this case, the plasmapause boundary may move outward past geosynchronous orbit and the increase will be due to measurements being made inside the plasmapause.  A second possible reason for the increase could be due to magnetospheric processes.

@@

Our first analysis uses data only from the time interval 1989-1991, which corresponds with the interval used in the analysis in \cite{Takahashi2010}. In this interval, there were 329 $D_{st}$ events.  For both types of events, the hour of the threshold crossing defines the zero epoch time, and for each event, $4\cdot24$ hours were considered before and $8\cdot24$ hours after each event.  To compute the epoch averages on a daily time scale, the median value of all available measurements for all events centered on a window of $\pm 12$ hours of the epoch zero hour was computed, and these averaging windows were shifted in increments of $24$~hours. Similar results are obtained if we first reduce each event time series to have a 1-day cadence by computing medians for each event in 1-day bins and then compute the medians across events.  Error bars for each parameter are the standard deviation of the values used in computing the median divided by the square root of the number of values used.

The stack plot in the upper panel of Figure \ref{DailyAveragedDstEvents} shows the epoch averages for the 329 events from 1989-1991.  The minimum $D_{st}$ median is $-48$~nT.  Consistent with \cite{Takahashi2010}, $D_{st}$ events correspond to elevated $\rho_{eq}$ before and after the event, although the magnitude of increase observed here is $1.5$ amu/cm$^3$ (from 18 to 19.5~amu/cm$^3$) instead of the increase of $\sim 10$ amu/cm$^3$ found in Figure 11 of \cite{Takahashi2010}.  The vertical green bars in the $\rho_{eq}$ plot show the number of $\rho_{eq}$ values that were used to compute the medians and red error lines.  For the non-\req\ medians, the number of measurements used in computing the medians is typically equal to the number of events, which is much larger than the number of \req\ values as we used all available measurements instead of restricting to only values where a non--fill $\rho_{eq}$ value existed.

In the lower panel of Figure \ref{DailyAveragedDstEvents} it is shown that the trends seen in the analysis of 1989-1991 also hold for the entire span of measurements from GOES 6, which spans 1983-1991 - a small elevation on the day of the event relative to the day before and day after, with median averages after the day of the event being slightly lower than that before.  
To determine if the magnitude of the density medians in the top panel of \ref{DailyAveragedDstEvents} on days -1, 0, and 1 are statistically different from each other, we used a two-population bootstrap test.  A bootstrap sample of the medians for each day was created by sampling the actual values with replacement.  The observed difference between the medians is shown in the table along with the fraction of observations in 1000 bootstrap differences with a different sign than that observed.  The results are shown in the fourth column of Table~\ref{BootstrapDifferenceTable}. The fourth column shows the results of a 2-population t-test and the fifth column indicates ???.

@@

We have also considered hourly averages as shown in the top panel of Fig.~\ref{HourlyAveragedDstEvents}.  A total of 668 events were found in the interval of May 1983 through August 1991 with an average duration of 9 hours and a median duration of 3 hours.  The error bars were computed in the same way as Fig.~\ref{DailyAveragedDstEvents}.  In this plot, the overall trend is similar to that found in the bottom panel of Fig.~\ref{HourlyAveragedDstEvents} - $\rho_{eq}$ is slightly elevated at the time of the threshold crossing of $D_{st}$ with respect to the values one day before and after, but the elevation is small relative to the error bars around the time of the event onset.  At this timescale no obvious change in mass density at hourly timescale is visible.

The bottom panel of Fig.~\ref{HourlyAveragedDstEvents} contains the events shown in the top panel under the constraint that $D_{st}$ remained below -50~nT for at least twelve hours after the event onset time.  In the time period of elevated mass density within 12 hours of the start of the event, there were an average of \inote{X} samples per hourly interval.  This result is consistent with previous observations of large mass density spikes in the plasma trough region during large geomagnetic storms (\cite{Yao2008}; \cite{Takahashi2010}.  To determine of this relationship can be observed in the events of the top panel of Fig.~\ref{HourlyAveragedDstEvents}, we sorted the events by the average $D_{st}$ values in the time interval of 0-12 hours, split the events in two and recomputed the epoch averages of $\rho_{eq}$.  The result is that there is no statistically significant difference in the mass density epoch averages for large versus small events.  This indicates that to confirm a statitical $\rho_{eq}$ dependence on large geomagnetic storms would require a dataset with a much larger number of strong geomagnetic storms.
